Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-01-21 boettiger.tex
writing

\subsubsection*{Arbitrary windows}\\
While superficially simple to use, the calculation of these statistics is subject to 
several arbitrary decisions and additional assumptions,
such as the choice of window size over which the statistic is computed,
the choice of detrending,
and whether and by how much the consecutive time-windows used in the calculation should overlap.  
\citet{Lenton2012} compare the performance of different approaches to these issues,
showing that choice of window and method used for detrending indeed influence the results,
and each approach considered has pros and cons.  
Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
it is difficult to evaluate and compare between such methods.  
A model-based approach side-steps these issue entirely.  
\cb{
 Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
 and that it assumes stationarity, which is precisely the thing we're looking to change?  
 Similarly could mention that the detrending approaches can remove the signal?
 We could also mention that some popular approaches such as autocorrelation require evenly-spaced points,
 and interpolating is incorrect, introducing fictitious correlations?
}

\subsubsection*{No quantitative measures}\\
The warning signal, ``an increase in statistic $x$,'' is not quantitative,
making it difficult to compare between signals or to
attribute a statistical significance to the detection.
Some authors have suggested Kendall's correlation coefficient to quantify an increase~\citep{Dakos2008, Dakos20011} 
in autocorrelation or variance.  
Other measures of increase, such as Pearson's, have also been proposed~\citep{Drake2010}.  
Much of the literature using summary statistics simply forgoes quantifying
the increase or estimating significance.  
While this may be satisfactory in experimental systems with controls and replicates to compare to,~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
any real-world application of early warning methods will not have the luxury of controls and replicates.
In such cases a quantitative definition of detection is essential.  
\cb{ We will see that correlation tests in summary statistics are an ineffective way to do this.  }

\subsubsection*{Problematic null models}\\
Specifying an appropriate null model is also difficult.  
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent, but this is
an assumption that is violated by the very experimental design:
since one variable is time and the other a moving time-averaged window.  
Under such a test any adequately large data set will find a significant result, regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
this destroys the natural autocorrelation in the time series.  
\cb{I feel citing \citet{Dakos2011} and \citet{Dakos2008} for these two examples would be impolite\dots} 
Parametric null models have been more promising, and have brought us closer to a model-based approach where assumptions are explicit.  
For instance, autoregressive models~\citet{Dakos2008} have been proposed, which are first parameterized from the data.  
\cb{is this next line okay to leave in?}
Estimating the null models directly from the data using a maximum likelihood framework, will provide the most powerful basis for this test.  
\citet{Seekell2011} proposes a conditional heteroscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}

@@

\subsubsection*{Hidden assumptions}\\
Assumes that the system sits effectively at equilibrium around a stable state  
and tests whether the system may be experiencing the slow change of a parameter that is weakening the stability of that state,
moving the system towards a saddle-node bifurcation.  
As such processes may be common in natural systems this is an important scenario to consider,
but without the clarity of assumptions made in a modeling approach there is a risk that
early warning signs will be used in situations where they do not apply.

Differing system dynamics
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in summary-statistic based approach.  
For instance, variance may increase for reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.   
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by 
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

\cb{ do we need the energy diagram figure to emphasize this is our assumption?}
Even restricting ourselves to this classic alternate stable state model,
we must still consider at least three possible scenarios that could explain a sudden transition.

%Differing causes of the transition.
Even when we can exclude or ignore such dynamics and 
restrict ourselves to systems that can support a saddle-node bifurcation,
any early warning signals approach must further assume that a changing parameter 
has brought the system closer to the bifurcation.  This excludes at least three alternative
explanations for the transition.  
The first is that a large perturbation of the system state 
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast. 
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.  
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.  
The third scenario is that the system does pass through a saddle-node bifurcation, 
but rather than gradually and monotonically approaching the critical point, the 
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

%%%% TOPIC FOR ANOTHER NOTE: Early warning signals and the Prosecutor's Fallacy %%%%%%%
%When applying detection approaches to data that have already experienced a critical transition,
%we have conditioned on the fact the transition has occurred.  This is a very different problem than 
%forecasting the probabilities of systems that have not collapsed!

and attempting to ascertain the cause of the transition
These are alternative null hypotheses 
 %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
\citep{Ditlevsen2010} % telling these two cases apart

5) Summary-statistic approaches inherently lack power. 
Methods for the detection of early warning signals are continually challenged by 
inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
After all, the challenge of extrapolating into the future across a dramatic shift is immense,
there is essentially no historical president for successfully \& reliably predicting regime shifts. % \cite Tetlock?
The signals one attempts to detect with these approaches are extremely weak.
Worse still, beyond this intuitive notion, there is little quantification of just how powerful our methods are,
even in artificial scenarios where uncertainy conforms to our parametric assumptions.  
% Seekell Am Nat, Contamin Eco Apps
We characterize the power

\subsection{Explicit use of the models}
Several recent approaches to early warning signals have considered fitting models, though have not considered the model-choice problem.  
Detection of warning signals as model choice (``non-parametric'' \citet{Carpenter2011e},  unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)

@@

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
% so this is not a reasonable statistical characterization
visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\citep{Dakos2008, Dakos2011, Dakos2009}.
Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
In principle, other quantifications of increasing variance, autocorrelation, etc.,
may perform better than $\tau$ and could be analyzed just as well in this framework. 
While we use this established metric of increase, we also introduce a novel warning indicator
that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supp. materials or main text?  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Models}
The detection of early warning is based on a \emph{local} theory,
predicting changes in the dynamics near the current attractor.
Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
but we can approximate the model near the bifurcation.  
Many possible complex models will correspond to this approximation.

%\textbf{Saddle-node bifurcation.}
One of the most common critical transitions modeled in ecological systems with alternative stable states is a saddle-node bifurcation~\citep[\{emph{e.g.}[]{Scheffer2001, Scheffer2009, Guttal2008a, VanNes2007, Biggs2009}.
The individual-based simulation follows the master equation, 
\begin{align}
  P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
  b_n &= \frac{e K x^2}{X^2 + h_t^2} \\ 
  d_n &= e X_t - a_t   
\end{align}
This model can experience a saddle-node bifurcation through increased mortality by increasing either $e$ or $a$. 
The SDE approximation holds for reasonable population sizes:
\begin{equation}
dX_t = \left( \frac{e K x^2}{X^2 + h_t^2} - e X_t - a_t\right) dt + \sigma \sqrt{ \frac{e K x^2}{X^2 + h_t^2} + e X_t + a_t} dB_t \label{ass}
\end{equation}
This parameterization is only an example, of course many others are possible~\citep{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
Under appropriate transformation, all can be written in the same normal form for he saddle-node bifurcation,
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
While in principle this model could be estimated directly from the data,
if all observations have been prior to the crash (\emph{i.e.} near the desirable equilibrium)
we have no data which is informative about the nonlinearity and such estimates will be unreliable.  
Fortunately the theory of early warning signals is a \emph{local} one, 
and we can linearize around the equilibrium to obtain a model we can estimate both more easily and more accurately.

Transforming the canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$,
which gives the model we refer to as the Linearized Saddle Node, LSN: 
\begin{equation}
\ud X = 2\sqrt{ r_t } (\phi - X_t)\ud t + \sigma\sqrt{\phi } \ud B_t. \label{LSN}
\end{equation}

\citet{Drake2010} induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} to test early warning signals.  
This bifurcation also arises in many common models.  For example, the stochastic version of~\citet{Levins1969}'s model (a logistic model),
\begin{equation}
\ud X_t = \left( c_t X_t (1-X_t/K) - e_t X_t \right) \ud t + \sigma \sqrt{\frac{e_t}{c_t}} \ud B_t \label{levins},
\end{equation}
where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent)
colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).  
The model contains a transcritical bifurcation when $c_t < e_t$.  The normal form of the bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t x - x^2 
\label{transcritical}
\end{equation}
We can express the linearized stochastic dynamics then by the time dependent mean-reverting process. 
We refer to this model as the linearized transcritical bifurcation, LTC.  
\begin{equation}
\ud X_t = r_t (r_t - X_t/K) \ud t + \sigma \sqrt{1+r_t} \ud B_t \label{LTC}
\end{equation}

%The pattern of the saddle-node differs from that of the transcritical in the square-root dependence on the bifurcation parameter.
Note that the transcritical bifurcation involves a gradual decline towards zero,
whereas the saddle-node bifurcation involves a sudden transition from a non-trivial equilibrium point.
Thus saddle-node bifurcations are the more typical focus of early warning efforts, even though both exhibit the phenomenon of critical slowing down.

Likelihood calculations also require specifying the rate of change of $r_t$, 
which we take to first order as
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power. When the system is stable, $r_t$ is constant and both models~\eqref{LSN} and~\eqref{LTC} reduce to a simple Ornstein-Uhlenbeck process, 
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval. For the linearized transcritical:

\begin{align}
\frac{\ud }{\ud t} E(x_i| M = \text{LTC})&=  r(t)(r(t) - x_i) \\
\frac{\ud}{\ud t} V(x_i| M = \text{LTC}) &=  -2 r(t) V(x_i|M) + (1+r(t))\sigma^2 
\label{LTCsoln}
\end{align}

For the linearized saddle-node bifurcation:
\begin{align}
\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
\label{LSNsoln}
\end{align}

These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supp. materials or main text?  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\subsubsection*{Arbitrary windows}
While superficially simple to use, the calculation of these statistics is subject to 
several arbitrary decisions and additional assumptions,
such as the choice of window size over which the statistic is computed,
the choice of detrending,
and whether and by how much the consecutive time-windows used in the calculation should overlap.  
\citet{Lenton2012} compare the performance of different approaches to these issues,
showing that choice of window and method used for detrending indeed influence the results,
and each approach considered has pros and cons.  
Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
it is difficult to evaluate and compare between such methods.  
A model-based approach side-steps these issue entirely.  
\cb{
 Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
 and that it assumes stationarity, which is precisely the thing we're looking to change?  
 Similarly could mention that the detrending approaches can remove the signal?
 We could also mention that some popular approaches such as autocorrelation require evenly-spaced points,
 and interpolating is incorrect, introducing fictitious correlations?
}

\subsubsection*{No quantitative measures}
The warning signal, ``an increase in statistic $x$,'' is not quantitative,
making it difficult to compare between signals or to
attribute a statistical significance to the detection.
Some authors have suggested Kendall's correlation coefficient to quantify an increase~\citep{Dakos2008, Dakos20011} 
in autocorrelation or variance.  
Other measures of increase, such as Pearson's, have also been proposed~\citep{Drake2010}.  
Much of the literature using summary statistics simply forgoes quantifying
the increase or estimating significance.  
While this may be satisfactory in experimental systems with controls and replicates to compare to,~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
any real-world application of early warning methods will not have the luxury of controls and replicates.
In such cases a quantitative definition of detection is essential.  
\cb{ We will see that correlation tests in summary statistics are an ineffective way to do this.  }

\subsubsection*{Problematic null models}
Specifying an appropriate null model is also difficult.  
Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
assumes only that the two variables are independent, but this is
an assumption that is violated by the very experimental design:
since one variable is time and the other a moving time-averaged window.  
Under such a test any adequately large data set will find a significant result, regardless of whether a warning signal exists.
A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
this destroys the natural autocorrelation in the time series.  
\cb{I feel citing \citet{Dakos2011} and \citet{Dakos2008} for these two examples would be impolite\dots} 
Parametric null models have been more promising, and have brought us closer to a model-based approach where assumptions are explicit.  
For instance, autoregressive models~\citet{Dakos2008} have been proposed, which are first parameterized from the data.  
\cb{is this next line okay to leave in?}
Estimating the null models directly from the data using a maximum likelihood framework, will provide the most powerful basis for this test.  
\citet{Seekell2011} proposes a conditional heteroscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}

@@

\subsubsection*{Hidden assumptions}
Assumes that the system sits effectively at equilibrium around a stable state  
and tests whether the system may be experiencing the slow change of a parameter that is weakening the stability of that state,
moving the system towards a saddle-node bifurcation.  
As such processes may be common in natural systems this is an important scenario to consider,
but without the clarity of assumptions made in a modeling approach there is a risk that
early warning signs will be used in situations where they do not apply.

%Differing system dynamics
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in summary-statistic based approach.  
For instance, variance may increase for reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.   
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by 
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

\cb{ do we need the energy diagram figure to emphasize this is our assumption?}
%Differing causes of the transition.
Even when we can exclude or ignore such dynamics and 
restrict ourselves to systems that can support a saddle-node bifurcation,
any early warning signals approach must further assume that a changing parameter 
has brought the system closer to the bifurcation.  This excludes at least three alternative
explanations for the transition.  
The first is that a large perturbation of the system state 
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast. 
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}  
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.  
The third scenario is that the system does pass through a saddle-node bifurcation, 
but rather than gradually and monotonically approaching the critical point, the 
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection{Summary-statistic approaches have less statistical power.}
Methods for the detection of early warning signals are continually challenged by 
inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
The challenge of extrapolating into the future across a dramatic shift is immense;
there is essentially no historical president for successfully \& reliably predicting regime shifts
outside of controlled settings. % \cite Tetlock?
The signals one attempts to detect with these approaches are extremely weak.
Despite the widespread recognition of the this need for large data sets, 
there has been very few studies quantitative studies of power looking at how much data is actually required~\citep{Contamin2009}.
\cb{this is the only example that claims to measure statistical power.  I find it incredible that Ellison, contrary to his own ecological statistics textbook, decides to just invent an arbitrary statistic (log ratio of the maximum value in the statistic seen before and after onset of the perturbation) and call this a measure of ``statistical power.''}
The Neyman-Pearson Lemma guarantees us that the most powerful test between hypotheses compares the 
likelihood that the data was produced under each. This requires a model-based approach, which,
as we will see, can have substantially more power to detect early warning signals than summary-statistic based approaches.

\subsection{Explicit use of the models}
Model-based approaches are beginning to play a larger role in early warning signals, 
though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.  
While choosing appropriate models without system-specific knowledge is challenging, 
much can be accomplished by framing the implicit assumptions above into more precise equations.
\citet{Lade2011} introduces the idea of generalized models for early warning signals, and 
\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.  
While presented as a ``non-parametric'' approach,~\citet{Carpenter2011e} is essentially similar, where the dynamics are approximated 
by an explicit stochastic differential equation (SDE) whose coefficients are estimated from statistical moments of the data.

\subsection{Model Choice}
Representing

We introduce a similar set of SDE models to these, 
\cb{should this next part come later?}
and go on to illustrate how these models can be fit to data by maximum likelihood and compared 
by a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
characterizes the rate of missed detections and false alarms expected in the estimate 
(or equivalently, estimates both the power and statistical significance of the test).

\subsection{Models}
The normal form for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter, which may be slowly varying in time.  
Transforming this canonical form to allow for an arbitrary mean $\theta$,
the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi(r_t)$.
We expand around the fixed point and express in terms of the stochastic dynamics:

\begin{equation}
\ud X = \sqrt{ r_t } (\phi(r_t) - X_t)\ud t + \sigma\sqrt{\phi(r_t) } \ud B_t. \label{LSN}
\end{equation}

It is also possible to derive this same expression from the master equation of various stochastic birth/death (or gains \& losses) models,
that contain saddle-node bifurcations, as we show in the appendix.  
Alternate dynamical forms are possible but will not differ substantially.

As we discuss above, the early warning signal paradigm must also include some assumptions on how the bifurcation parameter, $r_t$, is changing. We assume a gradual, monotonic change which we approximate to first order: 
\begin{equation}
r_t = r_0 - m t.
\label{R_t}
\end{equation}
Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power. When the system is stable, $r_t$ is constant and the linearized saddle-node model~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process, 
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}
This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.

For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval. For the linearized saddle-node bifurcation:
\begin{align}
\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
\label{LSNsoln}
\end{align}

These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.

@@

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\citep{Dakos2008, Dakos2011, Dakos2009}.
Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
In principle, other quantifications of increasing variance, autocorrelation, etc.,
may perform better than $\tau$ and could be analyzed just as well in this framework. 
While we use this established metric of increase, we also introduce a novel warning indicator
that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).

@@
